Polynomial Time and Full Dependent Types

### 2023-07-11 paper/paper.tex
[ paper ] various writing

\documentclass[acmsmall,review]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}

@@

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  We combine dependent types for reasoning about programs with linear
  type systems for implicit polynomial time.
\end{abstract}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{type theory, implicit computational complexity, linear types}

@@

\subsection{Quantitative Type Theory}

@@

@@

  \inferrule*
  {\Gamma \vdash N \stackrel\sigma: M_1 =_S M_2}
  {\Gamma \vdash M_1 \equiv M_2 \stackrel\sigma: S}
\end{mathpar}
The equality type also has an $\eta$ rule demonstrating $\conRefl(M)$
as the only proof of equality \cite{hoffmann99dependent}.

@@

\begin{enumerate}
\item Non-deterministic polytime via a bit oracle -- needs a statement
  of success
\item Previous could also be interpreted as probabilistic polytime
  (access to coin flips); can we formalise some kind of cryptographic
  property this way? Have a look at hash functions in the
  Authenticated Data Structures paper.
\item Insertion sort, quicksort as polytime functions.
\item A type of polytime functions; via a reflection operator
\item Programming with explicit resource annotations, including
  reasoning about the resource requirements directly.
\end{enumerate}

@@

\newcommand{\cstaccess}{c_{\mathit{access}}}
\newcommand{\cstmkclo}{c_{\mathit{mkclo}}}
\newcommand{\cstapp}{c_{\mathit{app}}}
\newcommand{\cstmkpair}{c_{\mathit{mkpair}}}
\newcommand{\cstmkunit}{c_{\mathit{mkunit}}}
\newcommand{\cstTrue}{c_{\mathit{mktrue}}}
\newcommand{\cstFalse}{c_{\mathit{mkfalse}}}
\newcommand{\cstLetpair}{c_{\mathit{letpair}}}
\newcommand{\cstSeq}{c_{\mathit{seq}}}
\newcommand{\cstIf}{c_{\mathit{if}}}

@@

\bob{introductory paragraph.}

The Hofmann-Dal Lago approach to proving the soundness of implicit
computational complexity systems is based on a three way coupling
between abstract mathematical elements (the \emph{what}), values from
a machine model (the \emph{how}), and resource potentials (the
\emph{fuel}). Each type in the system is defined as a three way
relation between these elements. The set of abstract elements depends
on the type being interpreted (e.g., types of natural numbers will be
defined in terms of the set $\mathbb{N}$). The machine model is fixed
across all types. We describe the particular machine model we use for
this paper in \autoref{sec:machine-model}. Potentials are arranged
into \emph{resource monoids} that we define in
\autoref{sec:resource-monoids}. Following Hofmann and Dal Lago, we
will select resource monoids appropriate to the kind of system that we
want to prove complexity soundness for.

@@

Note: the Agda formalisation is intrinsically well-scoped, which
makes expressing things a bit easier.

The syntax and rules of our target language are given in
\autoref{fig:opsem}.

\begin{enumerate}
\item Justification that looking up variables is constant: Due to
  lexical scoping, the size of the maximum variable lookup is bounded
  for a fixed program.
\item An advantage of our more ``elementary'' operational semantics
  (rather than encoding everything in the $\lambda$-calculus)
\end{enumerate}

@@

Resource potentials are attached to values to represent the amount of
intrinsic potential they have to fuel computation. Resource potentials
are organised into resource monoids. To motivate the definition, we
first enumerate the structure we will. To be able to account for the
combined potential attached to composite data and programs (e.g.,
pairs, or functions applied to arguments) we will require monoid
structure on potentials. The action of turning potential difference
into fuel for computation will be modelled by a difference
function. Finally, we require that our resource monoid contains
sufficient elements to fuel constant time operations. We gather these
requirements into a formal definition as follows:

\begin{definition}
  A \emph{resource monoid} $M$ consists of:
  \begin{enumerate}
  \item A carrier set $|M|$, whose elements represent amounts of
    potential. We use Greek letters $\alpha$, $\beta$, $\gamma$ to
    denote elements of a resource monoid.
  \item Commutative monoid structure $(\rplus,\rzero)$ on $|M|$, so
    we can add potentials.
  \item A \emph{difference function} $M : |M| \times |M| \to \natinf$,
    where $\natinf$ is the natural numbers extended with a negative
    infinity $- \infty$ and $- \infty + k = -\infty$. A difference
    $M(\alpha, \beta) = k \in \mathbb{N}$ means that starting with
    potential $\alpha$ and ending with potential $\beta$ yields $k$
    units of fuel. A difference of $- \infty$ means that $\alpha$
    contains insufficient potential to reach $\beta$. Differencing
    must satisfy:
    \begin{enumerate}
    \item for all $\alpha$, $M(\alpha,\alpha) = 0$; and
    \item for all $\alpha, \beta, \gamma$,
      $M(\alpha, \beta) \rplus M(\beta, \gamma) \leq M(\alpha, \gamma)$.
    \end{enumerate}
    The latter is a ``reverse triangle inequality'': the fuel
    recoverable by moving between potential levels $\alpha$ and
    $\gamma$ via $\beta$ may be less than the fuel recoverable
    moving from $\alpha$ to $\gamma$ directly.
  \item \bob{Interaction between $(\rplus,\rzero)$ and $M(\alpha,\beta)$.}
  \item An \emph{accounting function}
    $\mathit{acct} : \mathbb{N} \to |M|$ such that for all $k$,
    $k \leq M(\mathit{acct}(k),\rzero)$.
  \end{enumerate}
\end{definition}

@@

\begin{definition}[Natural Number Resource Monoid]
  Monoid structure is given by normal addition. Differencing is
  defined as
  \begin{displaymath}
    \mathbb{N}(m,n) = \left\{
      \begin{array}{ll}
        m - n & m \geq n \\
        - \infty & \textrm{otherwise}
      \end{array}\right.
  \end{displaymath}
  and $\mathit{acct}(k) = k$. Note that this is the simplest possible
  resource monoid due to the requirement that the $\mathit{acct}$
  function exists. \bob{Forward ref to where this is used.}
\end{definition}

Every resource monoid induces a partial ordering on its carrier set by
$\alpha \leq \beta$ iff $0 \leq M(\alpha, \beta)$. In the case of the
natural number resource monoid, this yields the usual ordering on the
naturals.

@@

The differencing operator of the natural number resource monoid can
only supply as much fuel as is contained in the potential. This makes
it suitable for amortised analysis, where we directly store potential
in data structures. For the two polynomial time systems, we need more
sophisticated structures originally presented by Dal Lago and
Hofmann. The fundamental idea with both is to represent potentials as
pairs $(m,p)$, where $m$ is a natural number and $p$ is a
polynomial. The $m$ tracks the ``size'' of data as it pertains to the
number of times an operation will be repeated by iterating over it ---
for example, an iterable natural number will have size equal to
itself, but a non-iterable natural number may be assigned zero
size. The polynomial $p$ tracks the complexity of a program as a
function of the size of the input. This leads to a differencing
operator that evaluates the polynomial with the size of the data:

\begin{definition}[Polynomial Resource Monoids]
  The \emph{Max-Polynomial} resource monoid $\MaxPoly$ has carrier
  set consisting of pairs $(m,p)$ where $m$ is natural number and $p$
  is a polynomial with natural number coefficients. Addition of
  elements is defined as $(m,p) \oplus (n,q) = (m \sqcup n, p+q)$ with
  $\emptyset = (0,0)$. Difference is defined as:
  \begin{displaymath}
    \MaxPoly((m,p),(n,q)) = \left\{
      \begin{array}{ll}
        p(m) - q(m)&m \geq n \textrm{ and } \forall k \geq m. p(k) \geq q(k) \\
        - \infty & \textrm{otherwise}
      \end{array}
    \right.
  \end{displaymath}
  $\MaxPoly$ accounts for constant time with constant polynomials:
  $\mathit{acct}(k) = (0,\lambda x.k)$.

@@

It is perhaps easier to see how the differencing operator works in the
special case of the difference $\MaxPoly((m,p),(0,0)) = p(m)$. I.e.,
if we have code that contains data of size $m$ and a program with
complexity $p$, then running the combination with no expectation of
remaining potential yields $p(m)$ available steps.

The $\MaxPoly$ and $\PlusPoly$ resource monoids will be used for the
cons-free and LFPL-style (i.e., with diamonds) systems
respectively. We will elaborate on how in
\autoref{sec:realisability-model} and how they support the two kinds
of natural number iteration in \autoref{sec:realising-iteration}.

FIXME: difference between this definition and Hofmann-Dal Lago's: we
use polynomials with natural number coefficients instead of a function
that is bounded by a polynomial.

@@

For both $\MaxPoly$ and $\PlusPoly$, the elements with zero size
component, i.e., of the form $(0,p)$, form a resource sub-monoid that
we will use for interpreting programs. We will call these sub-monoids
$\MaxPoly_0$ and $\PlusPoly_0$. We make use of resource sub-monoids
in our definitions of the interpretations of simultaneous
substitutions (\autoref{defn:rl-morphism}) and terms
(\autoref{defn:rl-terms}).

\subsubsection{Resource Monoids as Enriched Categories}

\bob{move this to below the definition of resource monoid}

An alternative perspective on resource monoids is to see them as a
kind of entriched category \cite{kelly}. Enriched categories replace
the sets of morphisms between objects with objects from some category
other than $\Set$. Define the posetal category $\natinf$ with objects $\mathbb{N} \cup \{\infty\}$ and

$\natinf$ as a category.

\bob{remember to look at the Marsden and thingy paper about
  quantitative resources and enriched categories.}  A concise
definition of Hofmann and Dal Lago's resource monoids can be given as:
\begin{definition}
  A \emph{resource monoid} is a strict symmetric monoidal
  $\natinf$-enriched category.
\end{definition}

\subsection{Models of Quantitative Type Theory}
\label{sec:qtt-models}

\bob{PLAN: RCwFs, and state that they can be constructed from Indexed Linear Posets}

\cite{atkey18} describes a general class of QTT models termed
\emph{Quantitative Categories with Families} (QCwFs). This class of
models fits will with the syntax of QTT, but to actually build models
it is revealing to construct them from certain indexed partial orders,
as we explain now. In the next section, we will construct a class of
specific models that prove the soundness of our complexity constrained
systems.

\begin{definition}[Category with Families]
  BLAH
\end{definition}

\begin{definition}
  A \emph{linear preorder} is a preordered set $(A, \leq)$ with
  \begin{enumerate}
  \item A commutative monoid $(I, -\otimes-)$ that is monotone
    w.r.t. the order; and
  \item Is closed: there is an operation
    $\multimap : A \times A \to A$ such that $x \otimes y \leq z$ iff
    $x \leq y \multimap z$.
  \item Right adjoint
  \end{enumerate}
  The collection of all linear preorders and functions that preserve the
  order and the operations forms a category $\LinPreorder$.
\end{definition}

\begin{definition}
  \begin{enumerate}
  \item A CwF $(\cat{C}, \top, \langle-,-\rangle, ...)$ that supports
    $\Pi$ and $\Sigma$ types.
  \item A functor $L : \cat{C}^\op \to \LinPreorder$, such that
    reindexing along projections has a right adjoint:
    \begin{displaymath}
      L_{\Gamma.A}(\pi^*X, Y) \cong L_{\Gamma}(X, \forall_A Y)
    \end{displaymath}
  \end{enumerate}
\end{definition}

\bob{define the construction}
\begin{enumerate}
\item Start with a CwF $\cat{C}$ that supports $\Pi$, $\Sigma$ and
  natural numbers.
\item Assume a functor $L : \cat{C}^\top \to \LinPreorder$ with
  reindexing along projections having a right adjoint.
\item Construct a new CwF with the same base category as $\cat{C}$,
  but with $\Ty'(\Delta) = \Sigma A : \Ty(\Delta).~L(\Delta.A)$ and
  $\Tm'(\Delta, (T, X)) = \Tm(\Delta, T)$.
\item
\end{enumerate}

This construction is implicit in \cite{atkey2018qtt}, where the
indexed linear poset is defined from a $\mathcal{R}$-LCA.

@@

Equipped with our underlying costed model of computation
(\autoref{sec:machine-model}) and a compositional notion of resource
potential (\autoref{sec:resource-monoids}), we construct models of QTT
that witness the resource and type soundness of our complexity
constrained systems.

\bob{Tie back to previous section}

We fix a resource monoid $M$ with distinguished sub-monoid $M_0$.

\subsubsection{Base CwF}
\label{sec:realisability-base-cwf}

As our base CwF, we use the category $\Set$ of sets and functions with
the standard interpretation of $\Pi$, $\Sigma$ types and the
$\mathbb{N}$ type.

@@

$\mathbb{N}$-Graded exponentials in each $L(A)$ are defined using the
action $(\cdot)$ of $(\mathbb{N}, \leq)$ on $M$ defined above \bob{do
  this}. At the level of realising values, the modality $\oc_n$ has no
effect; it only serves to alter the required potentials.
\begin{displaymath}
  \oc_n\,X = \{(a,\alpha,v) \mid \exists \alpha'.\,M(n \cdot \alpha', \alpha) = 0 \land (a, \alpha', v) \in X \}
\end{displaymath}

@@

To complete the construction of $L$ as an indexed linear preorder, we
need to give realisers for each of the required inequalities in
\autoref{def:indexed-linear-preorder}. In each case, this is a matter
of programming in the language of \autoref{sec:machine-model}. The
corresponding potentials are calculated by counting the steps in the
ensuing programs. Note that, so far, all realisers are constant time
(relative to their input), so the constructions so far work for any
resource monoid. For systems that require iteration, we will need more
structure on our resource monoids \bob{fwd ref}.

\begin{proposition}
  $L$, with $I$, $\otimes$, $\multimap$, $\oc_n$, and $\forall_{B}$ defined above, is
  an indexed linear preorder.
\end{proposition}

\begin{proof}
  See the Agda development.
\end{proof}

\subsubsection{Non-iterable Data Types}

>>>>

\documentclass[acmsmall,review,screen]{acmart}
\settopmatter{printfolios=true,printccs=true,printacmref=true}

@@

\newcommand{\lemref}[1]{\hyperref[#1]{Lemma \ref*{#1}}}
\newcommand{\exref}[1]{\hyperref[#1]{Example \ref*{#1}}}
\newcommand{\judgeref}[1]{\hyperref[#1]{Judgement \ref*{#1}}}
\newcommand{\propref}[1]{\hyperref[#1]{Proposition \ref*{#1}}}
\newcommand{\diagref}[1]{\hyperref[#1]{Diagram \ref*{#1}}}
\newcommand{\thmref}[1]{\hyperref[#1]{Theorem \ref*{#1}}}
\newcommand{\defref}[1]{\hyperref[#1]{Definition \ref*{#1}}}

\newcommand{\AgdaModule}[1]{\textcolor{blue}{\tt #1}}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  We combine dependent types with linear type systems that soundly and
  completely capture polynomial time computation. To combine dependent
  and linear types, we use \emph{Quantitative Type Theory}. We explore
  two systems for capturing polynomial time: one system that disallows
  construction of iterable data, and one, based on the LFPL system of
  Martin Hofmann, that controls construction via a payment
  method. Both of these are extend to full dependent types, allowing
  for arbitrary computation in types alongside guaranteed polynomial
  time computation in terms.

  The resulting system smoothly combines reasoning with dependent
  types and restricted computational complexity. \bob{Elaborate after
    Section 3 is written.}We prove the soundness of the systems using
  a realisability technique due to Hofmann and Dal Lago.
\end{abstract}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{type theory, implicit computational complexity, linear logic, computational complexity}

@@

\subsection{Quantitative Type Theory}
\label{sec:qtt-basic}

@@

\bob{Usage weakening somewhere}

@@

  \inferrule*
  {\Gamma \vdash N \stackrel0: M_1 =_S M_2}
  {\Gamma \vdash M_1 \equiv M_2 \stackrel0: S}
\end{mathpar}
The equality type also has an $\eta$ rule demonstrating $\conRefl(M)$
as the only proof of equality \cite{hoffmann99dependent}. Note that
equality reflection only targets the $\sigma = 0$ fragment, we cannot
use equality reflection to convey realisability information.

@@

% \begin{enumerate}
% \item Non-deterministic polytime via a bit oracle -- needs a statement
%   of success
% \item Previous could also be interpreted as probabilistic polytime
%   (access to coin flips); can we formalise some kind of cryptographic
%   property this way? Have a look at hash functions in the
%   Authenticated Data Structures paper.
% \item Insertion sort, quicksort as polytime functions.
% \item A type of polytime functions; via a reflection operator
% \item Programming with explicit resource annotations, including
%   reasoning about the resource requirements directly.
% \end{enumerate}

\subsection{Building Datatypes}
\label{sec:prog-datatypes}

We have only defined an iterable natural number datatype for both of
our systems above. Using the power of dependent types with a universe
we can use these natural numbers to create further datatypes whose
size is measured by a natural. Iteration on the size yields iteration
over the full datastructure. For example, in the LFPL system, we can
define a type of iterable lists by pairing a size with a type of
elements defined by recursion:
\begin{displaymath}
  \mathrm{IList}(A) = \Sigma n \stackrel1: \tyNat.\,\mathrm{rec}_{x.\mathsf{U}}\,n\,\{ \conZero(d) \mapsto I; \conSucc(d,n;p) \mapsto A \otimes p \}
\end{displaymath}

Nil and Cons constructors.

List iterator:

\bob{U notation for the universe above}

\begin{enumerate}
\item Insertion sort in LFPL.
\item Insertion sort in LFPL with correctness specification.
\end{enumerate}

\subsection{Polytime Problems}

Uses

\begin{displaymath}
  (A : \mathsf{U}, P : A \to \mathsf{U})
\end{displaymath}

Cartesian container morphisms

\subsection{Polytime-based Complexity Classes}

Monads

\subsubsection{Non-deterministic Polynomial Time}

\subsubsection{Probabilistic Polynomial Time}

@@

\newcommand{\cstaccess}{1}
\newcommand{\cstmkclo}{1}
\newcommand{\cstapp}{1}
\newcommand{\cstmkpair}{1}
\newcommand{\cstmkunit}{1}
\newcommand{\cstTrue}{1}
\newcommand{\cstFalse}{1}
\newcommand{\cstLetpair}{1}
\newcommand{\cstSeq}{1}
\newcommand{\cstIf}{1}

% \newcommand{\cstaccess}{c_{\mathit{access}}}
% \newcommand{\cstmkclo}{c_{\mathit{mkclo}}}
% \newcommand{\cstapp}{c_{\mathit{app}}}
% \newcommand{\cstmkpair}{c_{\mathit{mkpair}}}
% \newcommand{\cstmkunit}{c_{\mathit{mkunit}}}
% \newcommand{\cstTrue}{c_{\mathit{mktrue}}}
% \newcommand{\cstFalse}{c_{\mathit{mkfalse}}}
% \newcommand{\cstLetpair}{c_{\mathit{letpair}}}
% \newcommand{\cstSeq}{c_{\mathit{seq}}}
% \newcommand{\cstIf}{c_{\mathit{if}}}

@@

In this section and the next, we establish the polytime soundness of
our extensions of QTT by adapting a realisability method due to
\cite{hofmann-dallago}. This approach is based on a three way coupling
between abstract mathematical elements (the \emph{what}), values from
a machine model (the \emph{how}), and resource potentials (the
\emph{fuel}). Each type in the system is defined as a three way
relation between these elements. The set of abstract elements depends
on the type being interpreted (e.g., types of natural numbers will be
defined in terms of the set $\mathbb{N}$). The machine model is fixed
across all types. We describe the particular machine model we use for
this paper in \autoref{sec:machine-model}. Potentials are arranged
into \emph{resource monoids} that we define in
\autoref{sec:resource-monoids}. Following Hofmann and Dal Lago, we
will select resource monoids appropriate to the kind of system that we
want to prove complexity soundness for.

\paragraph{Agda Formalisation} The key soundness results in this
section have been formalised in the Agda proof assistant
\cite{agda}. The Agda formalisation can be found in the supplementary
material. After each definition and result we provide a pointer to the
Agda modules where the corresponding formalisation can be found, and
note interesting features of the mechanisation.

@@

We demonstrate that every program that can be written in extensions of
QTT has the complexity bounds that we claim by translating QTT terms
into an untyped CBV $\lambda$-calculus with a costed operational
semantics. The syntax and rules of our target language are given in
\autoref{fig:opsem}.

Variables are represented as de Bruijn indicies $i, j$. Expressions
$E \in \ExpSet$ can be (anonymous) $\lambda$-abstractions, unit,
pairing and boolean values, variables, sequencing, application, pair
elimination, and conditionals. Note that, with the exception of
$\lambda$-abstraction and sequencing, expressions never contain nested
expressions; instead referring to variables already defined. Values
$V \in \ValSet$ can be closures $\clo{E}{\eta}$, where $\eta$ is an
environment for the closure, unit values, pairs and booleans.

Costed evaluation of expressions in environments is defined by a
big-step operational semantics $E, \eta \Downarrow_k V$, where $k$ is
the number of steps. For simplicity, all operations cost $1$ unit,
though this could be generalised to allow for different operations to
have different costs. We use $\eta[i]$ to access the $i$th variable in
the environment, counting from the right. The evaluation rules are
mostly as one would expect, except for the application rule which
includes a self reference to the closure being invoked, in order to
allow recursive definitions.

\paragraph{Agda Formalisation} The machine model is defined in the
Agda module \AgdaModule{MachineModel}. We use an intrinsicaly
well-scoped syntax, which ensures that all variable look up operations
are always well defined.

@@

As we mentioned above, resource potentials are attached to values to
represent the amount of intrinsic potential they have to fuel
computation. Resource potentials are organised into resource
monoids. To be able to account for the combined potential attached to
composite data and programs (e.g., pairs, or functions applied to
arguments) we will require monoid structure on potentials. The action
of turning potential difference into fuel for computation will be
modelled by a difference function. Finally, we require that our
resource monoid contains sufficient elements to fuel constant time
operations. We gather these requirements into a formal definition as
follows, which is a reformulation of the resource monoids of
\cite{hoffman-dallago}:

\begin{definition}
  A \emph{resource monoid} $M$ consists of:
  \begin{enumerate}
  \item A carrier set $|M|$, whose elements represent amounts of
    potential. We use Greek letters $\alpha$, $\beta$, $\gamma$ to
    denote elements of a resource monoid.
  \item Commutative monoid structure $(\rplus,\rzero)$ on $|M|$, so
    we can add potentials.
  \item a \emph{difference function} $M : |M| \times |M| \to \natinf$,
    where $\natinf$ is the natural numbers extended with a negative
    infinity $- \infty$ and $- \infty + k = -\infty$. A difference
    $M(\alpha, \beta) = k \in \mathbb{N}$ means that starting with
    potential $\alpha$ and ending with potential $\beta$ yields $k$
    units of fuel. A difference of $- \infty$ means that $\alpha$
    contains insufficient potential to reach $\beta$. Differencing
    must satisfy:
    \begin{enumerate}
    \item for all $\alpha$, $M(\alpha,\alpha) = 0$; and
    \item for all $\alpha, \beta, \gamma$,
      $M(\alpha, \beta) + M(\beta, \gamma) \leq M(\alpha, \gamma)$.
    \end{enumerate}
    The latter is a ``reverse triangle inequality'': the fuel
    recoverable by moving between potential levels $\alpha$ and
    $\gamma$ via $\beta$ may be less than the fuel recoverable
    moving from $\alpha$ to $\gamma$ directly.
  \item Differencing and the commutative monoid structure must satisfy:
    \begin{enumerate}
    \item $M(\alpha, \beta) \leq M(\alpha \rplus \gamma, \beta \rplus \gamma)$; and
    \item $M(\alpha, \rzero) = 0$.
    \end{enumerate}
  \item An \emph{accounting function}
    $\mathit{acct} : \mathbb{N} \to |M|$ such that for all $k$,
    $k \leq M(\mathit{acct}(k),\rzero)$.
  \end{enumerate}
\end{definition}

For any resource monoid $M$, we can define an action of $\mathbb{N}$
on $M$ as $n \cdot \alpha = \alpha \oplus \cdots \oplus \alpha$, where
the right-hand side has $n$ summands.

\paragraph{Alternative definition} Every resource monoid induces a
partial ordering on its carrier set by $\alpha \leq \beta$ iff
$0 \leq M(\alpha, \beta)$. Taking this idea further, we can
reformulate a resource monoid as a symmetric monoidal category
enriched in the symmetric monoidal category $\natinf$, where the
monoid structure is addition. Isomorphism of objects $\alpha, \beta$
in such a category amounts to having $M(\alpha, \beta) = 0$. The
conditions in the definition above amount to the usual identity and
composition laws for enriched categories.

\paragraph{Agda Formalisation} Resource monoids are defined in the
module \AgdaModule{Algebra.ResourceMonoid}. We use a formulation
closer to the enriched category theory definition for the actual
formalisation, because it avoids having to treat equality in the
monoid structure separately from the induced preorder on elements.

@@

\begin{definition}[Natural Number Resource Monoid]
  Monoid structure is given by normal addition. Differencing is
  defined as
  \begin{displaymath}
    \mathbb{N}(m,n) = \left\{
      \begin{array}{ll}
        m - n & m \geq n \\
        - \infty & \textrm{otherwise}
      \end{array}\right.
  \end{displaymath}
  and $\mathit{acct}(k) = k$. Note that this is the simplest possible
  resource monoid due to the requirement that the $\mathit{acct}$
  function must exist.
\end{definition}

@@

The differencing operator of the natural number resource monoid can
only supply as much fuel as is contained in the potential.
% This makes
% it suitable for amortised analysis, where we directly store potential
% in data structures.
For the two polynomial time systems, we need more sophisticated
structures, both originally presented by Dal Lago and Hofmann. The
fundamental idea with both is to represent potentials as pairs
$(m,p)$, where $m$ is a natural number and $p$ is a polynomial. The
$m$ tracks the ``size'' of data as it pertains to the number of times
an operation will be repeated by iterating over it --- for example, an
iterable natural number will have size equal to itself, but a
non-iterable natural number may be assigned zero size. The polynomial
$p$ tracks the complexity of a program as a function of the size of
the input. This leads to a differencing operator that evaluates the
polynomial with the size of the data:

\begin{definition}[Polynomial Resource Monoids]\label{def:polynomial-resource-monoids}
  The \emph{Max-Polynomial} resource monoid $\MaxPoly$ has carrier set
  consisting of pairs $(m,p)$ where $m$ is natural number and $p$ is a
  polynomial with natural number coefficients. Addition of elements is
  defined as $(m,p) \oplus (n,q) = (m \sqcup n, p+q)$, where $\sqcup$
  is the max operator, with $\emptyset = (0,0)$. Difference is defined
  as:
  \begin{displaymath}
    \MaxPoly((m,p),(n,q)) = \left\{
      \begin{array}{ll}
        p(m) - q(m)&m \geq n \textrm{ and } \forall k \geq m. p(k) \geq q(k) \\
        - \infty & \textrm{otherwise}
      \end{array}
    \right.
  \end{displaymath}
  $\MaxPoly$ accounts for constant time with constant polynomials:
  $\mathit{acct}(k) = (0,\lambda x.k)$.

@@

It is perhaps easier to see how the differencing operator works in the
special case of the difference $\MaxPoly((m,p),(0,0)) = p(m)$. I.e.,
if we have code that contains data of size $m$ and a program with
complexity $p$, then running the combination with no expectation of
remaining potential yields $p(m)$ available steps. The $\MaxPoly$ and
$\PlusPoly$ resource monoids will be used for the \emph{Cons-free} and
LFPL-style systems respectively, as we explain in
\autoref{sec:realising-iteration} and show how these resource monoids
yield the required polytime bounds on programs.

\paragraph{Agda Formalisation} The natural number resource monoid is
defined in the Agda module \AgdaModule{Algebra.ResourceMonoid.Nat} and the
 polynomial monoids are both defined in
\AgdaModule{Algebra.ResourceMonoid.Polynomial}. The definition is
parameterised by the ``size monoid'' operation (either $\sqcup$ or
$+$) used to compose sizes.

% FIXME: difference between this definition and Hofmann-Dal Lago's: we
% use polynomials with natural number coefficients instead of a function
% that is bounded by a polynomial.

@@

For both $\MaxPoly$ and $\PlusPoly$, the elements with zero size
component, i.e., of the form $(0,p)$, form a resource sub-monoid that
we will use for interpreting programs. We will call these sub-monoids
$\MaxPoly_0$ and $\PlusPoly_0$.

\subsection{Models of Quantitative Type Theory from Indexed Preorders}
\label{sec:qtt-models}

\cite{atkey18qtt} described a general class of QTT models termed
\emph{Quantitative Categories with Families} (QCwFs). This class of
models fits well with the syntax of QTT, but to actually build models
it is convenient to construct them from certain indexed linear
preorders. The BCI-realisability model construction in
\cite{atkey18qtt} is a special case of this construction, but most of
the details are the same, so we only sketch a special case here to
concentrate on the specifics of complexity bounded realisability.

\begin{definition}\label{def:indexed-linear-preorder}
  A \emph{$\mathbb{N}$-linear preorder} is a preordered set
  $(L, \leq)$ with
  \begin{enumerate}
  \item a commutative monoid $(I, -\otimes-)$ that is monotone
    w.r.t. the order;
  \item is closed: there is an operation
    $\multimap : L \times L \to L$ such that $x \otimes y \leq z$ iff
    $x \leq y \multimap z$; and
  \item has a function $\oc : \mathbb{N} \to L \to L$, to interpret
    resource requirement adjustments, which satisfies the following
    inequalities:
    \begin{enumerate}
    \item $\oc_0 X \simeq I$, for discarding;
    \item $\oc_{m+n} X \leq (\oc_m X) \otimes (\oc_n X)$, for duplication;
    \item $\oc_m\oc_n X \leq \oc_{mn} X$ for nesting;
    \item $\oc_1 X \leq X$ for extraction / dereliction;
    \item $(\oc_n X) \otimes (\oc_n Y) \leq \oc_n(X \otimes Y)$, for distribution; and
    \item $n \leq m$ implies $\oc_n X \leq \oc_m X$, for usage weakening.
    \end{enumerate}
  \end{enumerate}
  The collection of all linear preorders and functions that preserve the
  order and the operations forms a category $\LinPreorder$.

  An \emph{indexed linear preorder} $L : \Set^\op \to \LinPreorder$ is
  a contravariant function, where we write $f^* : L(B) \to L(A)$ for
  the action of $L$ on functions $f : A \to B$, such that such that
  reindexing along projections has a right adjoint
  $L_{\Sigma_{a \in A}.B}(\pi_1^*X, Y) \cong L_{A}(X, \forall_B Y)$
  that commutes with reindexing.
\end{definition}

Given an indexed linear preorder $L : \Set^\op \to \LinPreorder$, we
construct a QCwF model of QTT with the basic type formers from
\autoref{sec:qtt-basic}:
\begin{enumerate}
\item Define a category $\cat{L}$ of interpretations of contexts with
  objects that are pairs $(A \in \Set, X \in L(A))$ and morphisms
  $f : (A,X) \to (B, Y)$ that are functions $f : A \to B$ such that
  $X \leq f^*Y$ (this is the Grothendieck category of $L$). There is a
  faithful functor $U : \cat{L} \to \Set$. The category $\cat{L}$ will
  be used for interpreting contexts in the $\sigma = 1$ fragment of
  QTT.
\item Define scaling of objects of $\cat{L}$ by
  $\pi(A,X) = (A, \oc_\pi X)$, and addition of $(A, X)$ and $(A, Y)$
  as $(A, X \otimes Y)$.
\item For each set $A$, define the collection of semantics types
  $\Ty(A)$ as the collection of $B : A \to \Set$ and
  $X \in L(\Sigma_{a \in A}.B(a))$. Thus a QTT type consists of an
  extensional meaning $B$ and its realisability specification $X$.
\item For each $A$ and $(B,X) \in \Ty(A)$, the $\sigma = 0$ fragment
  terms $\Tm(A,(B,X))$ are functions $\Pi_{a \in A}.\,B(a)$. For each
  context interpretation $(A, X)$ in $\cat{L}$ and type intepretation
  $(B,Y) \in \Ty(A)$, the $\sigma = 1$ fragment terms
  $\RTm((A,X),(B,Y))$ are functions $f : \Pi_{a \in A}.\,B(a)$ such
  that $X \leq \overline{f}^*Y$, where
  $\overline{f} : A \to \Sigma_{a \in A}.\,B(a)$ is the section
  associated with $f$.
\item The empty context is interpreted as $(\{*\}, I)$ and context
  extension $(A,X).(B,Y)$ (i.e., comprehension) by
  $(\Sigma_{a \in A}.B(a), \pi_1^* X \otimes Y)$.
\item Given $(A,X) \in \Ty(C)$ and
  $(B,Y) \in \Ty(\Sigma_{c \in C}.A(c))$, $\Sigma$-types are
  interpreted similarly to context extension and $\Pi$-types are
  interpreted as
  $(\lambda c.~(\Pi_{a \in A(c)}.\,B(c,a)), \forall_{A}(X \multimap
  (\mathit{ev}\,f)^*Y))$, where
  $\mathit{ev}\,f : (\Sigma_{c \in C}.A(c)) \to (\Sigma_{c \in
    C}.\Sigma_{a \in A(c)}.B(c,a))$ is defined using application of
  $f$.
\item Universe and Equality types are interpreted as normal in $\Set$
  with the realisability component set to $I$ in both cases. Note that
  the universe of small types includes realisability information for
  each type.
\item Realisability reflection for a type $(B,X) \in \Ty(A)$ is
  intepreted as the type
  $(\lambda a. \{ b \in B(a) \mid I \leq (\lambda a. (a,b))^*X \},
  I)$. Thus the set-component of the type is restricted to the
  elements that are realisable, while the actual realisability
  component is the ``empty'' $I$ realisability specification.
\end{enumerate}

\paragraph{Agda Formalisation} The indexed linear preorders are
defined in the Agda module \AgdaModule{IndexedLinear}. We have not yet
completed a formalisation of the construction of a full model of QTT
from an indexed linear preorder, nor have we constructed the syntax of
QTT in Agda, so these parts are currently unmechanised.

@@

Equipped with our underlying costed model of computation
(\autoref{sec:machine-model}) and a compositional notion of resource
potential (\autoref{sec:resource-monoids}), we can construct models of
QTT that witness the resource and type soundness of our complexity
constrained systems. We fix a resource monoid $M$ with distinguished
sub-monoid $M_0$.

@@

$\mathbb{N}$-Graded exponentials in each $L(A)$ are defined using the
action $(\cdot)$ of $(\mathbb{N}, \leq)$ on $M$ defined above. At the
level of the realising values, the modality $\oc_n$ has no effect; it
only serves to alter the required potentials.
\begin{displaymath}
  \oc_n\,X = \{(a,\alpha,v) \mid \exists \alpha'.\,M(n \cdot \alpha', \alpha) = 0 \land (a, \alpha', v) \in X \}
\end{displaymath}

@@

To complete the construction of $L$ as an indexed linear preorder, we
need to give realisers for each of the required inequalities in
\defref{def:indexed-linear-preorder}. In each case, this is a matter
of programming in the language of \autoref{sec:machine-model}. For
example, transitivity of the order is realised by sequencing of
realising expressions. The corresponding potentials are calculated by
counting the steps in the ensuing programs.

\begin{proposition}\label{prop:realisability-model}
  $L$, with $I$, $\otimes$, $\multimap$, $\oc_n$, and $\forall_{B}$
  defined above, is an indexed linear preorder.
\end{proposition}

\paragraph{Agda Formalisation} The construction of this indexed linear
preorder and the proof of \propref{prop:realisability-model} are
formalised in the Adga module
\AgdaModule{AmortisedRealisabilityModel}.

\subsubsection{Non-iterable Data Types}
\label{sec:noniter-model}

The model of QTT constructed in \propref{prop:realisability-model}
does not yet include any useful base types. Handling iterable types,
which are the ones that induce non-constant-time complexities,
requires specific properties of resource monoids that we introduce in
\autoref{sec:realising-iteration}.
